Off the beaten path: \\ Gaia reveals GD-1 stars outside of the main stream

### 2018-05-27 paper/submit-aas/referee-response.txt

We thank the referee for a constructive report that has improved the quality of
this work. Below, we have copied the remarks from the referee (indented by >)
and have responded to each point in text that immediately follows.

> The analysis of the Gaia DR2 data requires some care, especially at the faint
> end. For example, the authors do not discuss the amplitude of the correlations
> between the proper motion components and how these may affect their selection
> of tentative members. (By the way, the plots show relatively large ranges,
> particularly those of the proper motions and it would be useful if a few
> zoom-ins are included, so that the reader can assess better the data.).
>
> It would also be useful if the authors would explore how many stars in the
> regions around the stream have 2 parameter instead of 5 parameter solutions
> (i.e. they have position measured on the sky but no proper motions).
>
> Although the authors comment that all the stars have 8 or more visibility
> periods, this does not reflect completeness of the dataset but merely the
> reliability of the astrometric solution. Therefore maps showing the quality of
> the astrometric data in this region of the sky should be included as they
> enhance the credibility of the various features the authors claim to find. As
> a guide to what are the relevant diagnostics, the authors may wish to use
> Arenou et a. 2018 or Lindegren et al 2018.
>
> The presence of an overdensity that could correspond to the location of the
> progenitor is not so clear from the plots shown in the paper. Again a zoom-in
> might help here.

* APW:
* TODO: final figure summarize data quality and dust
* Push back on Zoom-ins? -> plot them for the referee, don't add much
* TODO: send Ana file with extinction estimates

> The authors need to include a measurement of the proper motion in the
> equatorial reference frame, for example, at the location of the highest
> density of the stream. This is clearly missing from the paper, and would make
> the paper useful to other readers.

* APW: easy, but stupid

> The authors do not use the Gaia photometry, which although shallower is likely
> to be significantly more accurate. How do the CMDs look like in the Gaia
> bands?

* AB: will check CMD
* Respond with plot and comment
* Main sequence becomes flatter

> The authors do not re-derive a distance to the stream, is there a reason for
> this? It would perhaps be useful to overlay the isochrone on the Gaia data to
> try to determine the distance to GD-1 more accurately, and hence also the
> tangential velocities of the stars in the stream.

* AB: Punt to next paper: respond that distance not the focus

> It was not clear to me how the solar reflex motion was corrected via the
> equation given, where does this stem from?

We have added a sentence to clarify how this was done.

> The significance of the various features (the blob, the spur) needs to be
> quantified. For example, Fig.4 seems to show a small overdensity in the "off
> blob" region on the proper motion diagram but there is no mention of this.

* APW/AB: Define significance

> The authors state that the proximity of the features to the location of the
> large gaps suggests that they might be related. Please elaborate on how.

* AB: explain the reference to Yoon et al., be explicit about the possible
  connection to the gap

> The orbit shown in Fig.3 is not integrated sufficiently long to be able to
> establish whether the stream has crossed the disc. Please include a comparison
> to an orbit integrated for 6-8 Gyr.

The orbit shown in Figure 3 crosses the disk midplane (z=0) twice in the segment of the orbit shown. Given uncertainties in the Milky Way's gravitational potential, and uncertainties in the full-space motion of the stream progenitor, we don't think that including a 6-8 Gyr integration in the paper will be too meaningful. However, we have extended the integration time for the paper figure to 1 Gyr, and attach a figure with a 6 Gyr integration with this response for the referee to see.

* Comments from the AAS statistics editor:

> In Fig 2, the authors choose a parabolic model for the GD-1 stream with
> (arbitrary?) lower/upper boundaries. This band clearly includes some
> background stars, and may miss some stream stars around \phi_1=-30d. It would
> be nice to identify stream members by some statistical technique, either in
> the 2D sky or 4D sky+kinematics space.

> I suggest the authors consult the excellent summary of modern `spatial point
> processes' methodology with cookbook to the comprehensive 'spatstat' R/CRAN
> package, by A.Baddeley et al. 'Spatial Point Patterns: Methodology and
> Applications with R' (2015). For example, I suspect a good list of stream
> members could emerge from a Voronoi tessellation with truncation of tile
> areas. Other possibilities include: nearest neighbors within a distance
> threshold, nonparametric density estimation, Allard-Fraley feature detection
> (MLE on Voronoi tessellation), and Byers-Raftery feature detection (MLE
> mixture model on k-NN distribution).

* APW: simple statistical technique in 4D, punt

### 2018-05-28 paper/submit-aas/referee-response.txt

> The authors do not use the Gaia photometry, which although shallower is likely
> to be significantly more accurate. How do the CMDs look like in the Gaia
> bands?

* AB: will check CMD
* Respond with plot and comment
* Main sequence becomes flatter

> The authors do not re-derive a distance to the stream, is there a reason for
> this? It would perhaps be useful to overlay the isochrone on the Gaia data to
> try to determine the distance to GD-1 more accurately, and hence also the
> tangential velocities of the stars in the stream.

* AB: Punt to next paper: respond that distance not the focus
- we do lay an isochrone in fig 1, consistent with a previously reported distance and stellar population
- however, we note that fig 4 indicates a distance gradient, closer at trailing than leading tail, also consistent with the literature
- while re-deriving a distance to the stream using would be useful, it is rather involved and would need to be done along the stream, so we are leaving it for our next contribution where we will derive more general properties of GD-1, while focusing on the new features in this one

### 2018-05-31 paper/submit-aas/referee-response.txt

We thank the referee for a constructive report that has improved the quality of
this work. Below, we have copied the remarks from the referee (indented by >)
and have responded to each point in text that immediately follows.

> The analysis of the Gaia DR2 data requires some care, especially at the faint
> end. For example, the authors do not discuss the amplitude of the correlations
> between the proper motion components and how these may affect their selection
> of tentative members. (By the way, the plots show relatively large ranges,
> particularly those of the proper motions and it would be useful if a few
> zoom-ins are included, so that the reader can assess better the data.).
>
> It would also be useful if the authors would explore how many stars in the
> regions around the stream have 2 parameter instead of 5 parameter solutions
> (i.e. they have position measured on the sky but no proper motions).
>
> Although the authors comment that all the stars have 8 or more visibility
> periods, this does not reflect completeness of the dataset but merely the
> reliability of the astrometric solution. Therefore maps showing the quality of
> the astrometric data in this region of the sky should be included as they
> enhance the credibility of the various features the authors claim to find. As
> a guide to what are the relevant diagnostics, the authors may wish to use
> Arenou et a. 2018 or Lindegren et al 2018.
>
> The presence of an overdensity that could correspond to the location of the
> progenitor is not so clear from the plots shown in the paper. Again a zoom-in
> might help here.

* APW:
* TODO: final figure summarize data quality and dust
* Push back on Zoom-ins? -> plot them for the referee, don't add much

- Dust map done
- 2-parameter solution done
- Source density at phi1 = [-40, -20, 0], phi2 = 0 is ~8000 sources/deg^2 -- should be close to 100% complete according to Arenou.
https://arxiv.org/pdf/1804.09375.pdf
https://arxiv.org/pdf/1804.09366.pdf

### 2018-05-31 paper/submit-aas/referee-response.txt

> The authors need to include a measurement of the proper motion in the
> equatorial reference frame, for example, at the location of the highest
> density of the stream. This is clearly missing from the paper, and would make
> the paper useful to other readers.

* APW: easy, but stupid

> The authors do not use the Gaia photometry, which although shallower is likely
> to be significantly more accurate. How do the CMDs look like in the Gaia
> bands?

The referee is correct that the Gaia photometry is significantly more accurate than PanSTARRS, however, at faint magnitudes, the regime where GD-1 members are most easily separable from the Milky Way field stars, the Gaia BP-RP color is much more uncertain than the PanSTARRS g-i color.
As a part of the response, we attach a figure that compares Gaia (left) and PanSTARRS (right) color-magnitude diagrams of GD-1 stars, selected with -1<phi_2<1, -50<phi_1<-10, and our fiducial proper motion selection.
Typical photometric uncertainties are shown as a function of magnitude, centered on 0 color in both panels.
PanSTARRS photometry is systematics dominated in the entire range, and the uncertainty in both g and i bands is 0.01 mag (0.014 mag for g-i color). 
Uncertainties in Gaia's G band are smaller in the entire range, but BP-RP's vary from 0.001 at G~15 to 0.05 at G~20, approximately 5 times larger than PanSTARRS at comparably faint magnitudes.
This leads to GD-1's main sequence being much better defined in PanSTARRS color-magnitude diagram than in Gaia's, so we opt to retain PanSTARRS photometry for selection of likely GD-1 members.

Q: do we comment on using g vs G?

### 2018-06-03 paper/submit-aas/referee-response.txt

> The authors need to include a measurement of the proper motion in the
> equatorial reference frame, for example, at the location of the highest
> density of the stream. This is clearly missing from the paper, and would make
> the paper useful to other readers.

* APW: easy, but stupid

> The authors do not use the Gaia photometry, which although shallower is likely
> to be significantly more accurate. How do the CMDs look like in the Gaia
> bands?

The referee is correct that the Gaia photometry is more accurate than PanSTARRS, however, at faint magnitudes, the regime where GD-1 members are most easily separable from the Milky Way field stars, the Gaia BP-RP color is much more uncertain than the PanSTARRS g-i color.
As a part of the response, we attach a figure that compares Gaia (left) and PanSTARRS (right) color-magnitude diagrams of GD-1 stars, selected within -1<phi_2<1, -50<phi_1<-10, and our fiducial proper motion selection.
Typical photometric uncertainties are shown as a function of magnitude, centered on 0 color in both panels.
PanSTARRS photometry is systematics dominated in the entire range, and the uncertainty in both g and i bands is 0.01 mag (0.014 mag for g-i color). 
Uncertainties in Gaia's G band are smaller in the entire range, but BP-RP's vary from 0.001 at G~15 to 0.05 at G~20, approximately 5 times larger than PanSTARRS at comparably faint magnitudes.
This leads to GD-1's main sequence being much better defined in PanSTARRS color-magnitude diagram than in Gaia's, so we opt to retain PanSTARRS photometry for selection of likely GD-1 members.

@@

> The authors do not re-derive a distance to the stream, is there a reason for
> this? It would perhaps be useful to overlay the isochrone on the Gaia data to
> try to determine the distance to GD-1 more accurately, and hence also the
> tangential velocities of the stars in the stream.

An isochrone at a fiducial distance to the stream is already overlaid on the GD-1 color-magnitude diagram (cmd) in Figure 1, and used to define a cmd selection box, which is very useful for improving the selection of likely GD-1 members.
Since the focus of this paper is reporting on new features of GD-1 detected in a cleaner map of the stream, using literature distance to the stream as a means to improve member selection seems adequate.
Of course, more precise distances, and hence tangential velocities, would be especially useful for dynamical modeling of GD-1, but improving on the current measurements would require a more involved calculation, so we leave it for future work where we plan to comprehensively derive stream properties in all six dimensions of the phase space.

### 2018-06-13 paper/submit-aas/referee-response.txt

> The significance of the various features (the blob, the spur) needs to be
> quantified. For example, Fig.4 seems to show a small overdensity in the "off
> blob" region on the proper motion diagram but there is no mention of this.

* APW/AB: Define significance

> The authors state that the proximity of the features to the location of the
> large gaps suggests that they might be related. Please elaborate on how.

To clarify the possible association, we've changed this statement to:
---
Both the spur and the blob are close to large gaps in the stream, so a mechanism that displaces stars from their original orbits along the stream to orbits beyond the stream would explain both sets of features. Future observations of the full 3-D kinematics combined with detailed dynamical modeling of the GD-1 system will be able to test this scenario.
---

### 2018-06-13 paper/gd1.tex
updated figure
response: spur-gap association

\newcommand{\todo}[1]{{\color{red} TODO: #1}}

@@

\title{Off the beaten path: \\
       Gaia reveals GD-1 stars outside of the main stream}
% Gaia DR2 reveals stream stars...
% A first look at the GD-1 stellar stream with Gaia DR2

@@

We retrieve data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$ in fields along the GD-1 stream.
% , and select
% batches of stars that have small latitude in the heliocentric GD-1 coordinate
% system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
We convert the equatorial sky coordinates and proper motions %of the parallax-selected stars
to the GD-1 coordinate system \citep[$\phi_1, \phi_2$,][]{Koposov:2010}, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

@@

Stars from the spur and the blob are similar to those along the main stream track in the color-magnitude diagram and in proper motions, which confirms their association to GD-1.
The proximity of these off-stream features to large gaps in the stream is suggestive of the gaps being their origin, further implied by hints of relative velocity differences.
For example, the median proper motion of the blob is offset from the adjacent stream field by $\approx0.3$\,mas\,yr$^{-1}$ in the positive $\mu_{\phi_1}$ direction.
We defer detailed modeling of these features to infer their co-evolution with the stream for future work.

>>>>

\newcommand{\todo}[1]{{\color{red} TODO: #1}}
\newcommand{\changes}[1]{{\color{green} #1}}

@@

\title{Off the beaten path: \\
       Gaia reveals GD-1 stars outside of the main stream}
% Gaia DR2 reveals stream dmlocal/tex/stars...
% A first look at the GD-1 stellar stream with Gaia DR2

@@

We retrieve data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$ in fields along the GD-1 stream.
% , and select
% batches of stars that have small latitude in the heliocentric GD-1 coordinate
% system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
We convert the equatorial sky coordinates and proper motions %of the parallax-selected stars
to the GD-1 coordinate system \citep[$\phi_1, \phi_2$,][]{Koposov:2010}, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\changes{In detail, we project the solar velocity vector onto the tangent space at each star's sky position, convert to angular motion using the assumed distance, and subtract the solar component from each star's proper motion (this is done using the \package{astropy} and \package{gala} packages).}
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

@@

Stars from the spur and the blob are similar to those along the main stream track in the color-magnitude diagram and in proper motions, which confirms their likely association to GD-1.
Both the spur and the blob are close to large gaps in the stream, so a mechanism that displaces stars from their original orbits along the stream to orbits beyond the stream would explain both sets of features.
Future observations of the full 3-D kinematics combined with detailed dynamical modeling of the GD-1 system will be able to test this scenario.

% The proximity of these off-stream features to large gaps in the stream is suggestive of the gaps being their origin, further implied by hints of relative velocity differences.
% For example, the median proper motion of the blob is offset from the adjacent stream field by $\approx0.3$\,mas\,yr$^{-1}$ in the positive $\mu_{\phi_1}$ direction.
% We defer detailed modeling of these features to infer their co-evolution with the stream for future work.
